### 2011-06-23 docs/architecture/architecture.tex
move some particle stuff from architecture.  Define \maestro and \castro
use \maestro

\section{MAESTRO Basics}

MAESTRO models problems that are in tight hydrostatic equilibrium.
The fluid state is decomposed into a 1D radial base state that
describes the hydrostatic structure of the star or atmosphere, and a
2- or 3D Cartesian full state, that captures the departures from
hydrostatic equilibrium.  Two basic geometries are allowed.  A {\em
  plane-parallel} geometry assumes that the domain is thin compared to
the radius of curvature of the star, and therefore the 1D base state
is perfectly aligned with the Cartesian state.  A {\em spherical}
geometry is for modeling an entire star.  Here, the 1D base state is
not aligned with the Cartesian state.  Figure~\ref{fig:base_state}
shows these geometries.

\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{\archfigpath/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{\archfigpath/base_spherical}
\caption[MAESTRO geometries]{\label{fig:base_state} MAESTRO geometries, showing both the
  1D base state and the full Cartesian state.  (Left) For multi-level
  problems in planar geometry, we force a direct alignment between the
  radial array cell centers and the Cartesian grid cell centers by
  allowing the radial base state spacing to change with space and
  time.  (Right) For multi-level problems in spherical geometry, since
  there is no direct alignment between the radial array cell centers
  and the Cartesian grid cell centers, we choose to fix the radial
  base state spacing across levels. Figure taken
  from~\cite{multilevel}.}
\end{figure}

MAESTRO can use adaptive mesh refinement to focus resolution on
complex regions of flow.  For Cartesian/plane-parallel geometries, all
cells at the same height must be at the same level of refinement.
This restriction is to allow for the base state to directly align with
the Cartesian state everywhere.  For spherical geometries, there is no
such restriction (again, see Figure~\ref{fig:base_state}).

\section{MAESTRO Directory Structure}

\begin{figure}[t]
\centering
\includegraphics[width=5.0in]{\archfigpath/maestro_directory2}
\caption[MAESTRO directory structure]
{The basic MAESTRO directory structure}
\end{figure}

All the files needed to build MAESTRO are contained in the {\tt
  fParallel} directory structure.  Problems piece together various
MAESTRO directories, choosing a reaction network, equation of state,
and conductivity routine to build an executable.  Briefly, the
sub-directories are:
\begin{itemize}
\item {\tt boxlib/}

 BoxLib is a library for describing meshes consisting of a union
 of boxes.  The BoxLib modules define the basic datatypes used
 in MAESTRO.  BoxLib also provides the routines that handle the
 parallelization and I/O.

@@

 A collection of various EOSes for use with the code.  The two
 important ones for MAESTRO are {\tt helmeos/} and {\tt
 gamma\_law\_general/}.

@@

 Various reaction networks for MAESTRO problems.  In addition to
 providing a routine to evolve the nuclear species due to reactions,
 the networks also define the species that are advected by the code.

@@

 The main MAESTRO algorithm directory.  The individual problem directories
 lie under here, and all of the main MAESTRO source lives in the {\tt Source/}
 sub-directory.

@@

   Several sub-directories containing routines to generate
   1D initial conditions in hydrostatic equilibrium for 
   mapping onto the MAESTRO grid.  See \S~\ref{sec:initial_models}
   for details.

@@

   The main MAESTRO source.  Here you will find the driver routine, the
   advection routines, etc.  All MAESTRO problems will compile this
   source.

@@

   Each problem in MAESTRO gets it own sub-directory.  The {\tt
     GNUmakefile} in the problem directory includes the instructions
   on how to build the executable, including what modules in {\tt
     extern} are used.

   Any file that you place in a sub-directory here takes 
   precedence over a file of the same name in {\tt MAESTRO/}.
   This allows problems to have custom versions of the main
   MAESTRO routines (e.g.\ initial conditions via {\tt initdata.f90}.
   See \S~\ref{sec:makefile} for details on the build system).

@@

  Some simple scripts that are useful for building, running,
  maintaining MAESTRO.

@@

In this section we give an overview of MAESTRO, including some of the
standard problems, how to run the code, some basic runtime parameters,
and how to look at the output.

@@

Different problems in MAESTRO are contained in individual
sub-directories under {\tt MAESTRO/}.  The {\tt GNUmakefile}
in each problem directory lists the components of {\tt MAESTRO}
that are used to build the executable.  Full details of the
{\tt GNUmakefile} can be found in \S~\ref{sec:adding_problems}.

@@

The {\tt flame} problem models a planar laminar flame.  Initially cool
fuel and hot ash are put in pressure equilibrium.  Thermal diffusion
transfers heat to the fuel, driving a burning front.  This problem
does not use the traditional MAESTRO elliptic constraint, but
rather enforces $\nabla \cdot U = S$ (through the runtime parameter
{\tt do\_smallscale = T}) as discussed in \cite{SNe}.

@@

To build the MAESTRO executable, simply type `{\tt make}' in the
problem directory.  The executable will have a name like {\tt
  main.Linux.Intel.exe}, where the specific parts of the name depend
on the compilers and OS used.  MAESTRO can be built with support for
parallelism (using MPI or MPI+OpenMP), or compiled as a serial code.
The top portion of the {\tt GNUmakefile} can be edited to choose the
build options.  Setting {\tt MPI := t} will build the code to use MPI
for parallelization.  Setting {\tt OMP := t} will use OpenMP.  OpenMP
is used together with MPI, with MPI distributing the grids across the
processors and within a shared-memory node, OpenMP allows many cores
to operate on the same grid.  A final build option determines whether
to build the code with support for debugging.  Setting {\tt NDEBUG :=
  t} disables debugging features.

@@

\item {\tt restart} tells MAESTRO to restart from a checkpoint.  The
      value of this parameter should be the file number to restart from.
      For example, to restart from the checkpoint file {\tt chk00010},
      you would set {\tt restart = 10}.

@@

{\tt Amrvis} is an easy-to-use visualization tool developed at LBL for
2- and 3D datasets which can plot slices through 3D datasets as well
as volume-renderings.  It can also very easily extract 1D lines
through the dataset along any coordinate direction.  It is distributed
separately from the MAESTRO distribution.

@@

By default, MAESTRO outputs global diagnostics each timestep into a file
called {\tt maestro\_diag.out}.  This includes the maximum Mach number,
peak temperature, and peak nuclear energy generation rate.  Individual problems
can provide their own {\tt diag.f90} file to produce custom diagnostic
output.

@@

In addition to the problems described above which use the full
capabilities of MAESTRO, there are a number of unit tests that
exercise only specific components of the MAESTRO solvers.  The
tests have their own drivers (a custom {\tt varden.f90}) that
initialize only the data needed for the specific test and call
specific MAESTRO routines directly.

@@

  This test initializes a Gaussian temperature profile and calls
  the thermal diffusion routines in MAESTRO to evolve the state 
  considering only diffusion.  The driver estimates a timestep
  based on the explicit thermal diffusion timescale and loops
  over calls to the thermal diffusion solver.  A Gaussian remains
  Gaussian when diffusing, so an explicit error can be computed
  by comparing to the analytic solution.  This test is 
  described in \cite{xrb}.

@@

Different MAESTRO problems are defined in sub-directories under
{\tt MAESTRO/}.  To add a problem, start by creating a new
sub-directory---this is where you will compile your problem and
store all the problem-specific files.

@@

The next line sets the compiler to be used for compilation:
\begin{lstlisting}[language={[gnu]make},mathescape=false]
  COMP := Intel
\end{lstlisting}
The MAESTRO build system knows what options to use for various
compiler families.  The {\tt COMP} flag specifies which compiler to
use.  Allowed values include {\tt Intel}, {\tt gfortran}, {\tt PGI},
and {\tt PathScale}.  The specific details of these choices are
defined in the {\tt MAESTRO/mk/comps/} directory.

@@

A MAESTRO application is built from several packages (the
multigrid solver, an EOS, a reaction network, etc.).  The core
MAESTRO packages are always included, so a problem only needs
to define the EOS, reaction network, and conductivities to
use, as well as any extra, problem-specific files.  
\begin{lstlisting}[language={[gnu]make},mathescape=false]
EOS_DIR := extern/EOS/helmeos   
CONDUCTIVITY_DIR := extern/conductivity/constant
NETWORK_DIR := extern/networks/ignition

@@

The runtime parameters for the core MAESTRO algorithm are listed
in {\tt MAESTRO/\_parameters}.  That file is parsed at compile-time
by the {\tt MAESTRO/write\_probin.py} script
(along with any other parameter files listed in the {\tt PROBIN\_PARAMETERS}
variable in the {GNUmakefile}, see \S \ref{sec:makefile}).  The script
outputs the {\tt probin.f90} source file.  Each line
in the {\tt \_parameters} file has the form: 
\vskip 3mm
{\em parameter} \hskip 10em  {\em data-type} \hskip 10em  {\em value} 
\vskip 3mm
\noindent where {\em parameter} is the name of the runtime parameter,
{\em data-type} is one of \{{\tt character}, {\tt real}, {\tt
  integer}, {\tt logical}\}, and the {\em value} specifies the default
value for the runtime parameter.  Comments are indicated by a `{\tt
  \#}' character and a used to produce documentation about the
available runtime parameters.

@@

MAESTRO models subsonic flows that are in hydrostatic equilibrium.
The solution in MAESTRO is broken up into a 1D base state and the 2-
or 3D full state.  The job of the 1D base state in the algorithm is
to represent the hydrostatic structure.  The full, Cartesian state
carries the departures from hydrostatic equilibrium.  The underlying
formulation of the low Mach number equations assumes that the base
state is in hydrostatic equilibrium.  At the start of a simulation,
the initial model is read in and taken as the base state.  Therefore,
any initial model needs to already be in hydrostatic equilibrium.

The routines in {\tt MAESTRO/initial\_model/} prepare an initial model
for MAESTRO.  There are two types of routines here, the first type
take an existing 1D initial model produced somewhere else (e.g.\ a
1D stellar evolution code), and and map it onto a uniform grid, at
the desired resolution, using the equation of state in MAESTRO, and
using MAESTRO's discretization of hydrostatic equilibrium.  The second
type generate the initial model internally, by integrating the
condition of hydrostatic equilibrium together with a simplifying
assumption on the energy (e.g. isothermal or isentropic).  In
both cases hydrostatic equilibrium is enforced as:
\begin{equation}
\frac{p_{i+1} - p_i}{\Delta r} = \frac{1}{2} (\rho_i + \rho_{i+1})
g_{i+1/2}
\end{equation}
Here, $g_{i+1/2}$ is the edge-centered gravitational acceleration.

@@

MAESTRO initial models are read in by the {\tt extern/model\_parser}
routines.  This expects the initial model to contain a header giving
the number of variables and their names, followed by rows of data
giving the coordinate and data values at that coordinate.  The initial
model should contain the same species data (as mass fractions) as
defined in the {\tt network} module used by the MAESTRO problem.

@@

MAESTRO's gridding is handled by the BoxLib library.  The underlying
idea in BoxLib is to allow for adaptive mesh refinement (AMR)---different
regions of the domain can have different spatial resolutions.  The
domain is broken into boxes.  At the coarsest (base) level of
refinement, the entire computational domain is covered, either by one
box or broken across many boxes.  Higher levels of refinement have
finer zones (typically $2\times$).  Only a portion of the domain may
be covered by the higher levels of refinement.

@@

On a grid, the data can be stored at cell-centers, on a face/edge, or
on the corners.  In BoxLib, data that is on an edge is termed `nodal'
in that direction (see Figure~\ref{fig:dataloc}).  Data that is on the
corners is nodal in all spatial directions.  In MAESTRO, the state
data (density, enthalpy, velocity, $\ldots$) is generally
cell-centered.  Fluxes are nodal in the direction they represent.
A few quantities are nodal in all directions (e.g.\ $\phi$ used in
the final velocity projection).

@@

A \fab\ would represent the data for a single box in the domain.
In MAESTRO, we don't usually deal with \fab s alone, but rather
we deal with \multifab s, described next.

@@

A \multifab\ is a collection of \fab s at the same level of
refinement.  This is the primary data structure that MAESTRO
routine operate on.  A multilevel simulation stores the 
data in an array of \multifab s, where the array index refers
to the refinement level.

@@

To build a \multifab, we need to provide a \layout, the number of
components to store in the \multifab\, and the number of ghostcells.  In
MAESTRO, the hierarchy of grids will be described by a single
\mllayout.  A \multifab\ can be declared and built at any time in a
simulation using the \mllayout, thereby allocating space at every
grid location in the simulation.  The sequence to build a \multifab\
appears as
\begin{lstlisting}[language={[95]fortran},mathescape=false]
  type(multifab) :: mfab(nlevs)
  ...
  do n = 1, nlevs
     call multifab_build(mfab(n), mla%la(n), nc, ng)
  enddo
\end{lstlisting}
Here, {\tt nc} is the number of components and {\tt ng} is the number
of ghostcells.  The \multifab\ is built one level at a time, using the
\layout\ for that level taken from the \mllayout, {\tt mla}.

@@

A \bctower\ holds the information about what boundary conditions are
in effect for each variable in a MAESTRO simulation.  \MarginPar{better description?}

\section{MAESTRO Data Organization}

The state of the star in MAESTRO is described by both a
multidimensional state and the 1D base state.  The full
multidimensional state is stored in \multifab s while the base state
is simply stored in Fortran arrays.  Here we describe the
major MAESTRO data-structures.

@@

\section{MAESTRO Helper Modules}

A number of MAESTRO modules appear frequently throughout the source.
Below, we describe some of the more common functionality of the most
popular modules.

@@

Presently there are 2 equations of state that work with MAESTRO:
\begin{itemize}
\item {\tt extern/EOS/helmeos/} represents a general stellar equation 
      of state, consisting of nuclei (as an ideal gas), radiation,
      and electrons (with arbitrary degeneracy and degree of relativity).
      This equation of state is that described in \cite{timmes_eos}.

@@

The {\tt eos\_module} declares for convenience the variables that need
appear in the {\tt eos} call argument list.  MAESTRO routines use
these module variables in the EOS call to avoid having to declare 
each quantity in each routine that calls the EOS.

@@

The {\tt network} module defines the number species advected by the
code ({\tt nspec}), their ordering, and gives their basic properties
(like atomic number, $Z$, and atomic mass, $A$).  All MAESTRO problems
require a {\tt network} module, even if there are no reactions
modeled.  Many different reaction modules (containing different sets
of isotopes) exist in {\tt extern/networks}.  The particular network
used by a problem is defined in the problem's {\tt GNUmakefile}.

@@

In MAESTRO, the state data is stored in a cell-centered multifab array
(the array index refers to the AMR level) and the MAC velocities are
stored in a 2D nodal multifab array (with indices referring to the AMR
level and the velocity component).  Here we demonstrate a typical way
to extract the state and MAC velocity data.

All MAESTRO routines are contained in a module, to allow for compile-time
argument checking.
\begin{lstlisting}[language={[95]fortran},mathescape=false]
module example_module

@@

\noindent Here, the {\tt bl\_types} and {\tt multifab\_module} modules
bring in the basic BoxLib data types. Specifically, here, {\tt
  bl\_types} defines {\tt dp\_t} which is the {\tt kind} used for
declaring double precision data, and {\tt multifab\_module} defines
the \multifab\ data type.  The {\tt ml\_layout\_module} defines the
datatype for a \mllayout---many routines will take an \mllayout
to allow us to fill ghostcells.  The {\tt variables} module is a MAESTRO
module that provides integer keys for indexing the state arrays.  In
this case the integer {\tt rho\_comp} refers to the location in the
state array corresponding to density.

@@

MAESTRO uses the multigrid solver to enforce the divergence constraint
both on the half-time edge-centered advective velocities (the ``MAC
projection'') and on the final cell-centered velocities (the ``HG
projection'').  For the MAC projection, since the velocity data is
edge-centered (the MAC grid), the projection is cell-centered.  For
the HG projection, since the velocity data is cell-centered, the
projection is node-centered.  \MarginPar{should draw a figure} The multigrid solver performs a number
of V-cycles until the residual drops by 10-12 orders-of-magnitude.
There are several options that affect how the multigrid solver
behaves, which we describe below.

@@

MAESTRO has support for Lagrangian particles that are passively
advected with the velocity field.  These are useful for diagnostics
and post-processing.  To use particles, particles must be seeded into
the domain by writing a problem-specific {\tt init\_particles.f90}
routine.  This routine is called at the start of the simulation.  The
{\tt init\_particles} routines add particles at specific locations by
calling the {\tt particle\_module}'s {\tt add} routine when a given
criteria is met by the fluid state.

@@

Some simple python scripts allow for the plotting of the particle
positions.  For example, running {\tt test2} with particles enabled
will seed particles in the initial hotspot.  To plot the results,
first set your {\tt PYTHONPATH} environment variable to point to the
{\tt fParallel/data\_processing/python/} directory, for example:
\begin{verbatim}
export PYTHONPATH="/home/username/development/MAESTRO/fParallel/data_processing/python"
\end{verbatim}
This will allow python to see the {\tt parseparticles.py} routine.
For the {\tt test2} problem, the {\tt plotparticles.py} routine shows
how to plot the particle histories and make an animation of the
particles colored by the ash mass fraction.  This script is run as:
\begin{verbatim}
./plotparticles.py timestamp_*
\end{verbatim}

Note, these python routines require the NumPy and matplotlib packages.
On a Fedora Linux system, the necessary routines can be installed via:
\begin{verbatim}
yum install python-matplotlib lyx-fonts stix-fonts
\end{verbatim}

>>>>

\section{\maestro\ Basics}

\maestro\ models problems that are in tight hydrostatic equilibrium.
The fluid state is decomposed into a 1D radial base state that
describes the hydrostatic structure of the star or atmosphere, and a
2- or 3D Cartesian full state, that captures the departures from
hydrostatic equilibrium.  Two basic geometries are allowed.  A {\em
  plane-parallel} geometry assumes that the domain is thin compared to
the radius of curvature of the star, and therefore the 1D base state
is perfectly aligned with the Cartesian state.  A {\em spherical}
geometry is for modeling an entire star.  Here, the 1D base state is
not aligned with the Cartesian state.  Figure~\ref{fig:base_state}
shows these geometries.

\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{\archfigpath/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{\archfigpath/base_spherical}
\caption[\maestro\ geometries]{\label{fig:base_state} \maestro\ geometries, showing both the
  1D base state and the full Cartesian state.  (Left) For multi-level
  problems in planar geometry, we force a direct alignment between the
  radial array cell centers and the Cartesian grid cell centers by
  allowing the radial base state spacing to change with space and
  time.  (Right) For multi-level problems in spherical geometry, since
  there is no direct alignment between the radial array cell centers
  and the Cartesian grid cell centers, we choose to fix the radial
  base state spacing across levels. Figure taken
  from~\cite{multilevel}.}
\end{figure}

\maestro\ can use adaptive mesh refinement to focus resolution on
complex regions of flow.  For Cartesian/plane-parallel geometries, all
cells at the same height must be at the same level of refinement.
This restriction is to allow for the base state to directly align with
the Cartesian state everywhere.  For spherical geometries, there is no
such restriction (again, see Figure~\ref{fig:base_state}).

\section{\maestro\ Directory Structure}

\begin{figure}[t]
\centering
\includegraphics[width=5.0in]{\archfigpath/maestro_directory2}
\caption[\maestro\ directory structure]
{The basic \maestro\ directory structure}
\end{figure}

All the files needed to build \maestro\ are contained in the {\tt
  fParallel} directory structure.  Problems piece together various
\maestro\ directories, choosing a reaction network, equation of state,
and conductivity routine to build an executable.  Briefly, the
sub-directories are:
\begin{itemize}
\item {\tt boxlib/}

 BoxLib is a library for describing meshes consisting of a union
 of boxes.  The BoxLib modules define the basic datatypes used
 in \maestro.  BoxLib also provides the routines that handle the
 parallelization and I/O.

@@

 A collection of various EOSes for use with the code.  The two
 important ones for \maestro\ are {\tt helmeos/} and {\tt
 gamma\_law\_general/}.

@@

 Various reaction networks for \maestro\ problems.  In addition to
 providing a routine to evolve the nuclear species due to reactions,
 the networks also define the species that are advected by the code.

@@

 The main \maestro\ algorithm directory.  The individual problem directories
 lie under here, and all of the main \maestro\ source lives in the {\tt Source/}
 sub-directory.

@@

   Several sub-directories containing routines to generate
   1D initial conditions in hydrostatic equilibrium for 
   mapping onto the \maestro\ grid.  See \S~\ref{sec:initial_models}
   for details.

@@

   The main \maestro\ source.  Here you will find the driver routine, the
   advection routines, etc.  All \maestro\ problems will compile this
   source.

@@

   Each problem in \maestro\ gets it own sub-directory.  The {\tt
     GNUmakefile} in the problem directory includes the instructions
   on how to build the executable, including what modules in {\tt
     extern} are used.

   Any file that you place in a sub-directory here takes 
   precedence over a file of the same name in {\tt MAESTRO/}.
   This allows problems to have custom versions of the main
   \maestro\ routines (e.g.\ initial conditions via {\tt initdata.f90}.
   See \S~\ref{sec:makefile} for details on the build system).

@@

  Some simple scripts that are useful for building, running,
  maintaining \maestro.

@@

In this section we give an overview of \maestro, including some of the
standard problems, how to run the code, some basic runtime parameters,
and how to look at the output.

@@

Different problems in \maestro\ are contained in individual
sub-directories under {\tt MAESTRO/}.  The {\tt GNUmakefile}
in each problem directory lists the components of {\tt MAESTRO}
that are used to build the executable.  Full details of the
{\tt GNUmakefile} can be found in \S~\ref{sec:adding_problems}.

@@

The {\tt flame} problem models a planar laminar flame.  Initially cool
fuel and hot ash are put in pressure equilibrium.  Thermal diffusion
transfers heat to the fuel, driving a burning front.  This problem
does not use the traditional \maestro\ elliptic constraint, but
rather enforces $\nabla \cdot U = S$ (through the runtime parameter
{\tt do\_smallscale = T}) as discussed in \cite{SNe}.

@@

To build the \maestro\ executable, simply type `{\tt make}' in the
problem directory.  The executable will have a name like {\tt
  main.Linux.Intel.exe}, where the specific parts of the name depend
on the compilers and OS used.  \maestro\ can be built with support for
parallelism (using MPI or MPI+OpenMP), or compiled as a serial code.
The top portion of the {\tt GNUmakefile} can be edited to choose the
build options.  Setting {\tt MPI := t} will build the code to use MPI
for parallelization.  Setting {\tt OMP := t} will use OpenMP.  OpenMP
is used together with MPI, with MPI distributing the grids across the
processors and within a shared-memory node, OpenMP allows many cores
to operate on the same grid.  A final build option determines whether
to build the code with support for debugging.  Setting {\tt NDEBUG :=
  t} disables debugging features.

@@

\item {\tt restart} tells \maestro\ to restart from a checkpoint.  The
      value of this parameter should be the file number to restart from.
      For example, to restart from the checkpoint file {\tt chk00010},
      you would set {\tt restart = 10}.

@@

{\tt Amrvis} is an easy-to-use visualization tool developed at LBL for
2- and 3D datasets which can plot slices through 3D datasets as well
as volume-renderings.  It can also very easily extract 1D lines
through the dataset along any coordinate direction.  It is distributed
separately from the \maestro\ distribution.

@@

By default, \maestro\ outputs global diagnostics each timestep into a file
called {\tt maestro\_diag.out}.  This includes the maximum Mach number,
peak temperature, and peak nuclear energy generation rate.  Individual problems
can provide their own {\tt diag.f90} file to produce custom diagnostic
output.

@@

In addition to the problems described above which use the full
capabilities of \maestro, there are a number of unit tests that
exercise only specific components of the \maestro\ solvers.  The
tests have their own drivers (a custom {\tt varden.f90}) that
initialize only the data needed for the specific test and call
specific \maestro\ routines directly.

@@

  This test initializes a Gaussian temperature profile and calls
  the thermal diffusion routines in \maestro\ to evolve the state 
  considering only diffusion.  The driver estimates a timestep
  based on the explicit thermal diffusion timescale and loops
  over calls to the thermal diffusion solver.  A Gaussian remains
  Gaussian when diffusing, so an explicit error can be computed
  by comparing to the analytic solution.  This test is 
  described in \cite{xrb}.

@@

Different \maestro\ problems are defined in sub-directories under
{\tt MAESTRO/}.  To add a problem, start by creating a new
sub-directory---this is where you will compile your problem and
store all the problem-specific files.

@@

The next line sets the compiler to be used for compilation:
\begin{lstlisting}[language={[gnu]make},mathescape=false]
  COMP := Intel
\end{lstlisting}
The \maestro\ build system knows what options to use for various
compiler families.  The {\tt COMP} flag specifies which compiler to
use.  Allowed values include {\tt Intel}, {\tt gfortran}, {\tt PGI},
and {\tt PathScale}.  The specific details of these choices are
defined in the {\tt MAESTRO/mk/comps/} directory.

@@

A \maestro\ application is built from several packages (the
multigrid solver, an EOS, a reaction network, etc.).  The core
\maestro\ packages are always included, so a problem only needs
to define the EOS, reaction network, and conductivities to
use, as well as any extra, problem-specific files.  
\begin{lstlisting}[language={[gnu]make},mathescape=false]
EOS_DIR := extern/EOS/helmeos   
CONDUCTIVITY_DIR := extern/conductivity/constant
NETWORK_DIR := extern/networks/ignition

@@

The runtime parameters for the core \maestro\ algorithm are listed
in {\tt MAESTRO/\_parameters}.  That file is parsed at compile-time
by the {\tt MAESTRO/write\_probin.py} script
(along with any other parameter files listed in the {\tt PROBIN\_PARAMETERS}
variable in the {GNUmakefile}, see \S \ref{sec:makefile}).  The script
outputs the {\tt probin.f90} source file.  Each line
in the {\tt \_parameters} file has the form: 
\vskip 3mm
{\em parameter} \hskip 10em  {\em data-type} \hskip 10em  {\em value} 
\vskip 3mm
\noindent where {\em parameter} is the name of the runtime parameter,
{\em data-type} is one of \{{\tt character}, {\tt real}, {\tt
  integer}, {\tt logical}\}, and the {\em value} specifies the default
value for the runtime parameter.  Comments are indicated by a `{\tt
  \#}' character and a used to produce documentation about the
available runtime parameters.

@@

\maestro\ models subsonic flows that are in hydrostatic equilibrium.
The solution in \maestro\ is broken up into a 1D base state and the 2-
or 3D full state.  The job of the 1D base state in the algorithm is
to represent the hydrostatic structure.  The full, Cartesian state
carries the departures from hydrostatic equilibrium.  The underlying
formulation of the low Mach number equations assumes that the base
state is in hydrostatic equilibrium.  At the start of a simulation,
the initial model is read in and taken as the base state.  Therefore,
any initial model needs to already be in hydrostatic equilibrium.

The routines in {\tt MAESTRO/initial\_model/} prepare an initial model
for \maestro.  There are two types of routines here, the first type
take an existing 1D initial model produced somewhere else (e.g.\ a
1D stellar evolution code), and and map it onto a uniform grid, at
the desired resolution, using the equation of state in \maestro, and
using \maestro's discretization of hydrostatic equilibrium.  The second
type generate the initial model internally, by integrating the
condition of hydrostatic equilibrium together with a simplifying
assumption on the energy (e.g. isothermal or isentropic).  In
both cases hydrostatic equilibrium is enforced as:
\begin{equation}
\frac{p_{i+1} - p_i}{\Delta r} = \frac{1}{2} (\rho_i + \rho_{i+1})
g_{i+1/2}
\end{equation}
Here, $g_{i+1/2}$ is the edge-centered gravitational acceleration.

@@

\maestro\ initial models are read in by the {\tt extern/model\_parser}
routines.  This expects the initial model to contain a header giving
the number of variables and their names, followed by rows of data
giving the coordinate and data values at that coordinate.  The initial
model should contain the same species data (as mass fractions) as
defined in the {\tt network} module used by the \maestro\ problem.

@@

\maestro's gridding is handled by the BoxLib library.  The underlying
idea in BoxLib is to allow for adaptive mesh refinement (AMR)---different
regions of the domain can have different spatial resolutions.  The
domain is broken into boxes.  At the coarsest (base) level of
refinement, the entire computational domain is covered, either by one
box or broken across many boxes.  Higher levels of refinement have
finer zones (typically $2\times$).  Only a portion of the domain may
be covered by the higher levels of refinement.

@@

On a grid, the data can be stored at cell-centers, on a face/edge, or
on the corners.  In BoxLib, data that is on an edge is termed `nodal'
in that direction (see Figure~\ref{fig:dataloc}).  Data that is on the
corners is nodal in all spatial directions.  In \maestro, the state
data (density, enthalpy, velocity, $\ldots$) is generally
cell-centered.  Fluxes are nodal in the direction they represent.
A few quantities are nodal in all directions (e.g.\ $\phi$ used in
the final velocity projection).

@@

A \fab\ would represent the data for a single box in the domain.
In \maestro, we don't usually deal with \fab s alone, but rather
we deal with \multifab s, described next.

@@

A \multifab\ is a collection of \fab s at the same level of
refinement.  This is the primary data structure that \maestro\
routine operate on.  A multilevel simulation stores the 
data in an array of \multifab s, where the array index refers
to the refinement level.

@@

To build a \multifab, we need to provide a \layout, the number of
components to store in the \multifab\, and the number of ghostcells.  In
\maestro\, the hierarchy of grids will be described by a single
\mllayout.  A \multifab\ can be declared and built at any time in a
simulation using the \mllayout, thereby allocating space at every
grid location in the simulation.  The sequence to build a \multifab\
appears as
\begin{lstlisting}[language={[95]fortran},mathescape=false]
  type(multifab) :: mfab(nlevs)
  ...
  do n = 1, nlevs
     call multifab_build(mfab(n), mla%la(n), nc, ng)
  enddo
\end{lstlisting}
Here, {\tt nc} is the number of components and {\tt ng} is the number
of ghostcells.  The \multifab\ is built one level at a time, using the
\layout\ for that level taken from the \mllayout, {\tt mla}.

@@

A \bctower\ holds the information about what boundary conditions are
in effect for each variable in a \maestro\ simulation.  \MarginPar{better description?}

\section{\maestro\ Data Organization}

The state of the star in \maestro\ is described by both a
multidimensional state and the 1D base state.  The full
multidimensional state is stored in \multifab s while the base state
is simply stored in Fortran arrays.  Here we describe the
major \maestro\ data-structures.

@@

\section{\maestro\ Helper Modules}

A number of \maestro\ modules appear frequently throughout the source.
Below, we describe some of the more common functionality of the most
popular modules.

@@

Presently there are 2 equations of state that work with \maestro:
\begin{itemize}
\item {\tt extern/EOS/helmeos/} represents a general stellar equation 
      of state, consisting of nuclei (as an ideal gas), radiation,
      and electrons (with arbitrary degeneracy and degree of relativity).
      This equation of state is that described in \cite{timmes_eos}.

@@

The {\tt eos\_module} declares for convenience the variables that need
appear in the {\tt eos} call argument list.  \maestro\ routines use
these module variables in the EOS call to avoid having to declare 
each quantity in each routine that calls the EOS.

@@

The {\tt network} module defines the number species advected by the
code ({\tt nspec}), their ordering, and gives their basic properties
(like atomic number, $Z$, and atomic mass, $A$).  All \maestro\ problems
require a {\tt network} module, even if there are no reactions
modeled.  Many different reaction modules (containing different sets
of isotopes) exist in {\tt extern/networks}.  The particular network
used by a problem is defined in the problem's {\tt GNUmakefile}.

@@

In \maestro, the state data is stored in a cell-centered multifab array
(the array index refers to the AMR level) and the MAC velocities are
stored in a 2D nodal multifab array (with indices referring to the AMR
level and the velocity component).  Here we demonstrate a typical way
to extract the state and MAC velocity data.

All \maestro\ routines are contained in a module, to allow for compile-time
argument checking.
\begin{lstlisting}[language={[95]fortran},mathescape=false]
module example_module

@@

\noindent Here, the {\tt bl\_types} and {\tt multifab\_module} modules
bring in the basic BoxLib data types. Specifically, here, {\tt
  bl\_types} defines {\tt dp\_t} which is the {\tt kind} used for
declaring double precision data, and {\tt multifab\_module} defines
the \multifab\ data type.  The {\tt ml\_layout\_module} defines the
datatype for a \mllayout---many routines will take an \mllayout
to allow us to fill ghostcells.  The {\tt variables} module is a \maestro\
module that provides integer keys for indexing the state arrays.  In
this case the integer {\tt rho\_comp} refers to the location in the
state array corresponding to density.

@@

\maestro\ uses the multigrid solver to enforce the divergence constraint
both on the half-time edge-centered advective velocities (the ``MAC
projection'') and on the final cell-centered velocities (the ``HG
projection'').  For the MAC projection, since the velocity data is
edge-centered (the MAC grid), the projection is cell-centered.  For
the HG projection, since the velocity data is cell-centered, the
projection is node-centered.  \MarginPar{should draw a figure} The multigrid solver performs a number
of V-cycles until the residual drops by 10-12 orders-of-magnitude.
There are several options that affect how the multigrid solver
behaves, which we describe below.

@@

\label{arch:sec:particles}

\maestro\ has support for Lagrangian particles that are passively
advected with the velocity field.  These are useful for diagnostics
and post-processing.  To use particles, particles must be seeded into
the domain by writing a problem-specific {\tt init\_particles.f90}
routine.  This routine is called at the start of the simulation.  The
{\tt init\_particles} routines add particles at specific locations by
calling the {\tt particle\_module}'s {\tt add} routine when a given
criteria is met by the fluid state.

@@

Some simple python scripts allow for the plotting of the particle
positions.  See \S~\ref{analysis:sec:particles} for details.
